---
title: "DATA 605 Assignment 13: Calculus"
author: "Dan Smilowitz"
date: "November 6, 2016"
output: 
  html_document: 
    highlight: tango
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = NA, fig.align = 'center')
```


## Programmed Solutions

### Derivative
The derivative of the equation $f(x) = x^3 + 2x^2$ can be calculated for a given value $a$ using
$$f'(a) = \lim_{x \to a} \frac{f(x) - f(a)}{x - a} = \lim_{x \to a} \frac{x^3 + 2x^2 - a^3 - 2a^2}{x-a}$$

To ease the computational stress of analytically calculating the limit by approaching $a$ in small steps, this equation can be simplified:

$$\begin{align}
f'(a) &= \lim_{x \to a} \frac{x^3 + 2x^2 - a^3 - 2a^2}{x-a} = \lim_{x \to a} \frac{(x - a)(x^2 + a^2 + 2x + 2a + ax)}{x - a} \\
&= \lim_{x \to a} \ x^2 + a^2 + 2x + 2a + ax
\end{align}$$

Since directly entering $x = a$ into this equation does not create division by zero, a fairly simple function can be coded to get the derivative for a given value of $a$:
```{r der-func}
fprime <- function(a) {
  x <- a
  x^2 + a^2 + 2*x + 2*a + a*x
}
```

This function can be tested for a number of values and compared to the results of the built-in `deriv` function:
```{r test-der}
a <- -3:3
fprime(a)

x <- a
as.vector(attr(eval(deriv(~ x^3 + 2*x^2, 'x')), 'gradient'))
```


### Integral
To calculated the definite integral of the function, the area under the function will be calculated:
$$I = \sum_{x=1}^{x=3} 3x^2 + 4x \ \Delta x$$
where $\Delta x = 1 \times 10^{-6}$.

```{r int}
del_x <- 1e-6
x_range <- seq(1, 3, del_x)
area <- 0
for (i in 2:length(x_range)) {
  y_high <- 3 * x_range[i]^2 + 4*x_range[i]
  y_low  <- 3 * x_range[i - 1]^2 + 4*x_range[i - 1]
  area_inc <- (y_high) * del_x
  area <- area + area_inc
}
area
```

This result of ~42 can be verified analytically, or using the built-in `integrate` function:
```{r test-int}
integrate(function(x) {3*x^2 + 4*x}, 1, 3)
```



## Analytical Solutions

### Problem 1
Setting $u = \sin(x)$ gives $du = \cos(x) dx$.  Rearranging this gives $du = \frac{dx}{\cos(x)}$.  Subtituting this into the original problem,
$$\int \sin(x) \cos(x) dx = \int u du = \frac{1}{2} u^2 + C$$
Now substituting back in for $u$,
$$\int \sin(x) \cos(x) dx = \frac{1}{2} \sin^2(x) + C$$


### Problem 2
$\int x^2 e^x dx$ can be represented as $\int u(x)dv$ where $u(x) = x^2$ and $dv = e^x dx$. Using these, $du = 2xdx$ and $v(x) = e^x$.

Using these functions to integrate by parts:  
$$\int u(x)dv = u(x)v(x) - \int v(x)du = x^2 e^x - 2 \int x e^x dx + C$$  

Performing integration by parts on the second integral, using $u(x) = x$ and $dv = e^x dx$; $du=dx$ and $v(x) = e^x$.
$$\int x^2 e^x dx = x^2 e^x - 2 \left[ x e^x - \int e^x dx \right] = x^2 - e^x - 2xe^x + C \\
\int x^2 e^x dx = e^x \left( x^2 - 2x +2 \right) + C$$


### Problem 3
For a product of functions, $\left[ f(x) g(x) \right]' = f'(x)g(x) + g'(x)f(x)$.

Substituting in $f(x) = x$ and $g(x) = \cos(x)$,

$$\frac{d}{dx} \left[ x \cos(x) \right] = 1 \times \cos(x) + (-\sin(x)) \times x = \cos(x) - x\sin(x)$$


### Problem 4
For an exponential function, $\frac{d}{dx} e^u = e^u \frac{du}{dx}$.  Using $u = 4x^3$,
$$\frac{d}{dx} e^{x^4} = e^{x^4} \times \frac{d}{dx} x^4 = 4 x^3 \times e^{x^4}$$
