---
title: "DATA 605 Assignment 2: Matrix Factorization"
author: "Dan Smilowitz"
date: "September 6, 2016"
output: 
  html_document: 
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = NA)
```


## Problem Set 1

### General Case
In general, for matrices $A$ and $B$,
$$AB \neq BA$$

Substituting in $B = A^T$ gives 
$$AA^T \neq A^TA$$


### Special Case
In the case where $A = B$, this equation becomes
$$AB = BA \longrightarrow AA = AA$$

$AA = AA$ is necessarily true; in this case,  $AB = BA$ is true.

Thus, the only situation under which $A^TA = AA^T$ is one where
$$A^T = A$$

This is true, by definition, for all **symmetic matrices**.


## Problem Set 2

The function below factorizes matrices into LU or LDU:

```{r factorize}
factorize <- function(A, LDU = FALSE) {
  if (nrow(A) == ncol(A)) {
    square <- TRUE
  } else {square <- FALSE}
  
  if (square == TRUE) {
    size <- nrow(A)
    L <- diag(size)
    U <- A
    
    for (i in 1:(size - 1)) {
      for (j in (i + 1):size) {
        # get multipliers
        L[j, i] <- U[j, i] / U[i, i]
        # pivots and multiplication
        U[j, ]  <- U[j, ] - L[j, i] * U[i, ]
      }
    }
    
    #results
    if (LDU == TRUE) {
      D <- diag(size)
      for (k in 1:size) {
        D[k, k] <- U[k, k]
        U[k, ] <- U[k, ] / D[k, k]
      }
      LU <- list("L" = L, "D" = D, "U'" = U)
    } else {
      LU <- list("L" = L, "U" = U)
    }
  } else {LU <- "Error: matrix is not square"}
  LU
}
```

### Testing

Testing with the example matrix $A$ from the lecture notes

$$A = \left[ \begin{array}{cc}1 & 1 & 2 \\2 & 1 & 0 \\3 & 1 & 1 \end{array} \right]$$

```{r test-LU}
A <- matrix(c(1, 2, 3, 1, 1, 1, 2, 0, 1), nrow=3)
factorize(A)
```

$$A = LU = \left[ \begin{array}{cc}1 & 0 & 0 \\2 & 1 & 0 \\3 & 2 & 1 \end{array} \right] \left[ \begin{array}{cc}1 & 1 & 2 \\0 & -1 & -4 \\0 & 0 & 3 \end{array} \right]$$

```{r test-LDU}
factorize(A, LDU = TRUE)
```

$$A = LDU' = \left[ \begin{array}{cc}1 & 0 & 0 \\2 & 1 & 0 \\3 & 2 & 1 \end{array} \right] \left[ \begin{array}{cc}1 & 0 & 0 \\0 & -1 & 0 \\0 & 0 & 3 \end{array} \right] \left[ \begin{array}{cc}1 & 1 & 2 \\0 & 1 & 4 \\0 & 0 & 1 \end{array} \right]$$
