---
title: "DATA 605 Assignment 14: Taylor Series"
author: "Dan Smilowitz"
date: "November 6, 2016"
output: 
  html_document: 
    highlight: tango
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = NA, fig.align = 'center')
library(ggplot2)
```

$$f(x) \approx \sum_{n=0}^{\infty} \frac{f^{(n)} (a)}{n!} (x-a)^n$$


## Equation 1

### Derivation

The first three derivatives for $f(x) = \frac{1}{1-x}$ are
$$f'(x) = \frac{1}{(1-x)^2} \\
f''(x) = \frac{2}{(1-x)^3} \\
f'''(x) = \frac{6}{(1-x)^4}$$

This can be generalized as
$$f^{(n)}(x) = \frac{n!}{(1-x)^{n+1}}$$

Substituting this into the general equation,
$$\frac{1}{1-x} \approx \sum_{n=0}^{\infty} \frac{n!}{(1-a)^{n+1}} \times \frac{(x-a)^n}{n!} = \frac{(x-a)^n}{(1-a)^n}$$

At $a=0$, this becomes
$$\frac{1}{1-x} \approx \sum_{n=0}^{\infty} x^n$$

#### Convergence
Using the ratio series for infinite series:
$$\frac{A_{n+1}}{A_n} = \frac{x^{n+1}}{x^n} = x \\
\lim_{n \to \infty} x = x$$

So the series converges for $|x| < 1$.

### Validation
The true values and series approximations are plotted for $n=25$:
```{r test-1}
taylor1 <- function(x, n) {
  sum <- 0
  for(i in 0:n) {
    sum <- sum + x^i
  }
  sum
}
```

```{r plot-1, echo=FALSE}
x1 <- seq(-0.99, 0.99, 0.01)
test1 <- data.frame(x = rep(x1, 2), y = c(1 / (1 - x1), taylor1(x1, 25)),
                    Type = c(rep('Equation', length(x1)), rep('Series', length(x1))))

ggplot(test1, aes(x, y, col = Type)) + geom_line() + scale_color_discrete('') +
  theme(legend.position='bottom') + ggtitle('Series Approximation for 1/(1-x)')
```

It can be seen in the plot above that there is some divergence between the function and its Taylor series approximation at the edges of the range (-1, 1).



## Equation 2

### Derivation

The derivative of $f(x) = e^x$ is $e^x$, so any derivative $f^{(n)}(x) = e^x$.

Substituting this into the general equation,

$$e^x \approx \sum_{n=0}^{\infty} e^x \times \frac{(x-a)^n}{n!} = \frac{e^x (x-a)^n}{n! (1-a)^n}$$

At $a=0$, this becomes
$$\frac{1}{1-x} \approx \sum_{n=0}^{\infty} \frac{x^n}{n!}$$

#### Convergence
Using the ratio series for infinite series:
$$\frac{A_{n+1}}{A_n} = \frac{x^{n+1}}{(n+1)!} \frac{n!}{x^n} = \frac{x}{n+1} \\
\lim_{n \to \infty} \frac{x}{n+1} = 0$$

So the series converges for all values of x.

### Validation
The true values and series approximations are plotted for $n=25$:
```{r test-2}
taylor2 <- function(x, n) {
  sum <- 0
  for(i in 0:n) {
    sum <- sum + x^i / factorial(i)
  }
  sum
}
```

```{r plot-2, echo=FALSE}
x2 <- seq(-5, 5, 0.05)
test2 <- data.frame(x = rep(x2, 2), y = c(exp(x2), taylor2(x2, 25)),
                    Type = c(rep('Equation', length(x2)), rep('Series', length(x2))))

ggplot(test2, aes(x, y, col = Type)) + geom_line() + scale_color_discrete('') +
  theme(legend.position='bottom') + ggtitle('Series Approximation for e^x')
```

It can be seen in the plot above that the function and its taylor series approximation entirely overlap.



## Equation 3

### Derivation

The first four derivatives for $f(x) = \ln(x)$ are
$$f^{(1)}(x) = \frac{1}{1+x} \\
f^{(2)}(x) = \frac{-1}{(1+x)^2} \\
f^{(3)}(x) = \frac{2}{(1+x)^3} \\
f^{(4)}(x) = \frac{-6}{(1+x)^4}$$

This can be generalized as
$$f^{(n)}(x) = (-1)^{n-1} \frac{(n-1)!}{(1+x)^{n}}$$

Substituting this into the general equation,
$$\ln(1+x) \approx \sum_{n=0}^{\infty} (-1)^{n-1} \frac{(n-1)!}{(1+a)^{n}} \times \frac{(x-a)^n}{n!} = (-1)^{n-1} \frac{(x-a)^n}{n (1+a)^n}$$

At $a=0$, this becomes
$$\ln(1+x) \approx \sum_{n=0}^{\infty} (-1)^{n-1} \frac{x^n}{n}$$

#### Convergence
Using the ratio series for infinite series:
$$\frac{A_{n+1}}{A_n} = - \frac{x^{n+1}}{n} \frac{n}{x^n} = - \frac{-xn}{n+1} \\
\lim_{n \to \infty} \frac{-xn}{n+1} = -x$$

So the series converges for $|x| < 1$.

### Validation
The true values and series approximations are plotted for $n=25$:
```{r test-3}
taylor3 <- function(x, n) {
  sum <- 0
  for(i in 1:n) { # starting at n=1 since n=0 causes Inf
    sum <- sum + (-1)^(i + 1) * x^i / i
  }
  sum
}
```

```{r plot-3, echo=FALSE}
x3 <- seq(-0.99, 0.99, 0.01)
test3 <- data.frame(x = rep(x3, 2), y = c(log(1 + x3), taylor3(x3, 25)),
                    Type = c(rep('Equation', length(x3)), rep('Series', length(x3))))

ggplot(test3, aes(x, y, col = Type)) + geom_line() + scale_color_discrete('') +
  theme(legend.position='bottom') + ggtitle('Series Approximation for ln(1+x)')
```

As in the first plot, there is some divergence between the function and its Taylor series approximation at the lower edge of the valid range near -1.
