---
title: 'DATA 605 Assignment 5: Least Squares'
author: "Dan Smilowitz"
date: "September 20, 2016"
output: 
  html_document: 
    highlight: tango
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = NA)
```


## Problem Set 1

$$A = \left[ \begin{array}{cc}1 & 0 \\1 & 1 \\1 & 3 \\1 & 4 \end{array} \right]\ ; \ \ \ b = \left[ \begin{array}{c}0 \\8 \\8 \\20 \end{array} \right]$$

```{r AtA-Atb}
A = matrix(c(1, 1, 1, 1, 0, 1, 3, 4), ncol = 2, byrow = FALSE)
b = matrix(c(0, 8, 8, 20), ncol = 1)

(AtA <- t(A) %*% A)
(Atb <- t(A) %*% b)
```

$$A^TA = \left[ \begin{array}{cc}4 & 8 \\8 & 26 \end{array} \right]\ ; \ \ \ A^Tb = \left[ \begin{array}{c}36 \\112 \end{array} \right]$$

```{r x-hat}
(x_hat <- solve(AtA, Atb))
```

$$\hat{x} = \left[ \begin{array}{c}1 \\4 \end{array} \right]$$

```{r e}
(e <- b - A %*% x_hat)
```

$$e = \left[ \begin{array}{c}-1 \\3 \\-5 \\3 \end{array} \right]\ ; \ \ \ ||e||^2 = `r sum(e^2)`$$

```{r p}
p <- matrix(c(1, 5, 13, 17), ncol = 1)
Atp <- t(A) %*% p
solve(AtA, Atp)
```

$$\hat{x_p} = \left[ \begin{array}{c}1 \\4 \end{array} \right]$$

```{r e-b-p}
identical(e, b - p)
```

```{r orthogonal}
sum(e * p)
for (i in 1:ncol(A)) {print(sum(e * A[, i]))}
```

The dot products of $e$ with $p$ and with the columns of $A$ yield zero; this shows that the vectors are orthogonal.



## Problem Set 2

The function below returns the inverse of a full-rank square matrix using co-factors.

```{r car-data}
# read in data and store in A & b
car_data <- read.table('https://raw.githubusercontent.com/dsmilo/DATA605/master/data/auto-mpg.data',
                       col.names = c("x1", "x2", "x3", "x4", "mpg"))
A <- as.matrix(car_data[, 1:4])
b <- as.matrix(car_data[, 5])

# find AtA & Atb
AtA <- t(A) %*% A
Atb <- t(A) %*% b

# find least squares solution
x_hat <- solve(AtA, Atb)

# find the error
e <- b - A %*% x_hat
e2 <- sum(e^2)
```

The solution vector returned by this script is
$$\hat{x} = \left[ \begin{array}{c}-0.030 \\0.157 \\-0.006 \\1.997 \end{array} \right]$$

That is to say that the least-squares estimate for $mpg$ is
$$\widehat{mpg} = -0.030 x_1 + 0.157 x_2 -0.006 x_3 + 1.997 x_4$$

The fitting error of this estimate, $||e||^2$, is `r as.character(round(e2, 2))`.
