---
title: 'DATA 605 Assignment 6: Random Variables'
author: "Dan Smilowitz"
date: "October 7, 2016"
output: 
  html_document: 
    highlight: tango
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = NA)
```


## Problem Set 1

### Two Dice
For each roll of the die, there are 6 possibilities, so
$$N_{3rolls} = 6 \times 6 \times 6 = 216$$

### Sum of 3
There are two possibilities for two rolls totalling 3: 1 then 2; and 2 then 1.  The total probability is 

$$\begin{align}
P(sum = 3) &= P(1, 2) + P(2, 1) \\
           &= P(1) \times P(2) + P(2) \times P(1) \\
           &= \frac{1}{6} \times \frac{1}{6} + \frac{1}{6} \times \frac{1}{6} \\
           &= \frac{2}{36} = \frac{1}{18} = `r round(1/18, 4)`
\end{align}$$

### Shared Birthday
The event that at least two people share a birthday is the complement of the event that no person in the room shares a birthday with anyone else in the room.  Imagining people are asked their birthdays sequentially, the second person has a $\frac{364}{365}$ chance of having a birthday different from the first person.  The third person then has a $\frac{363}{365}$ chance of not sharing a birthday with either of the first two people.  This continues on through the 25th person, who will have a $\frac{341}{365}$ chance of not sharing a birthday with the previous 24 people, assuming they all had distinct birthdays.  So the total probability is

$$\begin{align}
P(unique) &= \frac{364}{365} \times \frac{363}{365} \times \cdots \times \frac{342}{365} \times \frac{341}{365} \\ \\
          &= \frac{364 \times 363 \times \cdots \times 342 \times 341}{365^{24}} \\
          &\approx 0.4313 \\ \\ \\
P(shared) &= 1 - P(unique) = 0.5687
\end{align}$$

#### Larger Population

The probability of a shared birthday between $n$ people can be generalized as
$$P(shared)_n = 1 - \frac{364 \times \cdots \times (364 - n + 2)}{365^{n - 1}}$$

Coded into an R function,
```{r birthday}
same_bday <- function(n) {
  unique <- prod(364:(364 - n + 2)) / (365 ^ (n - 1))
  1 - unique
}
```

For 50 people, the chance of at least one shared birthday is $P(shared)_{50} = `r round(same_bday(50), 4)`$.


## Problem Set 2

### One-Word Probability
```{r word-prob}
word_prob <- function(flnm, ranked = FALSE, printAllWords = FALSE, returnTable = FALSE) {
  library(stringr)
  # read in file
  body <- read.delim(flnm, header = FALSE, stringsAsFactors = FALSE)
  # collapse and convert to lowercase
  body <- str_to_lower(paste(body, collapse = ' '))
  # remove anything that isnt a space or lowercase letter
  body <- str_replace_all(body, '[^[:lower:]^ ]', '')
  # convert to vector
  word_vec <- str_split(body, ' ')[[1]]
  # get count of words and convert to probability
  word_table <- as.data.frame(table(word_vec)[-1], stringsAsFactors = FALSE) #remove count of spaces
  suppressPackageStartupMessages(library(dplyr))
  word_table <- word_table %>% transmute(Word = word_vec, Probability = Freq / sum(Freq))
  if(ranked) {
    word_table <- arrange(word_table, desc(Probability))
  }
  print(head(word_table, ifelse(printAllWords, nrow(word_table), 10)))
  if(returnTable){return(word_table)}
}
```

```{r test-data}
word_prob('data/assign6.sample.txt', ranked = TRUE)
```

### Two-Word Probability
```{r two-words}
phrase_prob <- function(flnm, word1, word2) {
  # adapt previous function
  library(stringr)
  body <- read.delim(flnm, header = FALSE, stringsAsFactors = FALSE)
  body <- str_to_lower(paste(body, collapse = ' '))
  body <- str_replace_all(body, '[^[:lower:]^ ]', '')
  word_vec <- str_split(body, ' ')[[1]]
  word_table <- as.data.frame(table(word_vec)[-1], stringsAsFactors = FALSE)
  suppressPackageStartupMessages(library(dplyr))
  word_table <- word_table %>% transmute(Word = word_vec, Probability = Freq / sum(Freq))
  # get individual word probabilities
  prob_1 <- word_table[word_table$Word == word1, 2]
  prob_2 <- word_table[word_table$Word == word2, 2]
  # find count of words appearing together
  phrase_count <- str_count(body, paste(word1, word2)) + str_count(body, paste(word2, word1))
  # divide by the number of possible phrases -- 1 fewer than the number of words
  prob_phrase <- phrase_count[1] / (length(word_vec) - 1)
  probs <- data.frame(Term = c(word1, word2, paste0(paste(word1, word2), ";", paste(word2, word1))), 
                      Probability = c(prob_1, prob_2, prob_phrase), stringsAsFactors = FALSE)
  probs
}
```

```{r test-phrase}
phrase_prob('data/assign6.sample.txt', 'the', 'federal')
```
