---
title: 'DATA 605 Assignment 10: Page Rank'
author: "Dan Smilowitz"
date: "October 27, 2016"
output: 
  html_document: 
    highlight: tango
    theme: flatly
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = NA, fig.align = 'center')
```


### Matrices A & B
As shown in the notes, the matrix $A$ is given by
$$A = \left[ \begin{array}{c}
0 & \frac{1}{2} & \frac{1}{2} & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 \\
\frac{1}{3} & \frac{1}{3} & 0 & 0 & \frac{1}{3} & 0 \\
0 & 0 & 0 & 0 & \frac{1}{2} & \frac{1}{2} \\
0 & 0 & 0 & \frac{1}{2} & 0 & \frac{1}{2} \\
0 & 0 & 0 & 1 & 0 & 0
\end{array} \right]$$

Node 2 in this example is a *dangling node* i.e. it has no outlinks.  This means that the matrix $A$ above, as well as the decayed matrix $B$ would not be row-stochastic and would yield a page rank vector of nearly or entirely all-zero entries.  As such, the second row of $A$ is replaced with a uniform row vector with values of $/frac{1}{6}$, indicating that because there are no outlinks from node 2, there is an even probability that a user travels next to any of the six nodes in the example.  The revised matrix $A$ is given by

$$A = \left[ \begin{array}{c}
0 & \frac{1}{2} & \frac{1}{2} & 0 & 0 & 0 \\
\frac{1}{6} & \frac{1}{6} & \frac{1}{6} & \frac{1}{6} & \frac{1}{6} & \frac{1}{6} \\
\frac{1}{3} & \frac{1}{3} & 0 & 0 & \frac{1}{3} & 0 \\
0 & 0 & 0 & 0 & \frac{1}{2} & \frac{1}{2} \\
0 & 0 & 0 & \frac{1}{2} & 0 & \frac{1}{2} \\
0 & 0 & 0 & 1 & 0 & 0
\end{array} \right]$$

The matrix $B$ is obtained by
$$B = 0.85 \times A + \frac{0.15}{n} \approx \left[ \begin{array}{c}
0.025 & 0.45 & 0.45 & 0.025 & 0.025 & 0.025 \\
0.1667 & 0.1667 & 0.1667 & 0.1667 & 0.1667 & 0.1667 \\
0.3083 & 0.3083 & 0.025 & 0.025 & 0.3083 & 0.025 \\
0.025 & 0.025 & 0.025 & 0.025 & 0.45 & 0.45 \\
0.025 & 0.025 & 0.025 & 0.45 & 0.025 & 0.45 \\
0.025 & 0.025 & 0.025 & 0.875 & 0.025 & 0.025
\end{array} \right]$$

In R, these are stored as below:
```{r A&B}
A <- matrix(
  c(0, 1/2, 1/2, 0, 0, 0,
  0, 0, 0, 0, 0, 0,
  1/3, 1/3, 0, 0, 1/3, 0,
  0, 0, 0, 0, 1/2, 1/2,
  0, 0, 0, 1/2, 0, 1/2,
  0, 0, 0, 1, 0, 0),
  nrow = 6, byrow = TRUE)

A[2, ] <- rep(1/6, 6)

B <- 0.85 * A + 0.15 / nrow(A)
```


### Power Iterations
The following function is created to perform power iterations on $B$ until convergence, utilizing a uniform rank vector $$r^T = \left[ \begin{array}{c}
\frac{1}{6} & \frac{1}{6} & \frac{1}{6} & \frac{1}{6} & \frac{1}{6} & \frac{1}{6} \end{array} \right]$$

```{r power-iteration}
power_iterate <- function(mat, vec) {
  converged <- FALSE
  n <- 0
  while(!converged) {
    vec <- crossprod(mat, vec)
    n <- n + 1
    if(identical(crossprod(mat, vec), vec)) {
      converged <- TRUE
    }
  }
  print(paste('Converged in', n, 'iterations'))
  return(vec)
}

r <- matrix(rep(1/nrow(B), nrow(B)), ncol=1)
it_results <- power_iterate(B, r)
```

The page rank vector and associated page rankings, as calculated, are:
`r kable(data.frame(page = seq(1, 6), vector = round(it_results, 4), rank = rank(1 - it_results)), padding=0, align = 'c')`

### Eiegen-Decomposition
For this exercise, we are interested in the largest eigenvalue of $B$, as well as the associated *left* eigenvector of $B$.  Taking only the real components, this can be shown to be
```{r eigen}
Re(eigen(B)$values[1])
ev_results <- matrix(Re(eigen(t(B))$vectors[,1]))
```

The eigenvectors returned by the `eigen` function return normal vectors (i.e. vectors with length 1); the returned vector is divided by its sum to give a vector with a sum 1.  This vector is identical to the vector returned using the `power_iterate` function:
```{r un-norm}
ev_results <- ev_results / sum(ev_results)
identical(round(ev_results, 13), round(it_results, 13))
```

`r kable(data.frame(page = seq(1, 6), vector = round(ev_results, 4), rank = rank(1 - ev_results)), padding=0, align = 'c')`


### Visualizing the Network
Using the `igraph` package, the network can be visualized in a directed graph, and the page rank of the nodes in the network returned.
```{r graph, echo=FALSE}
library(igraph)
G <- graph(c(1, 2, 1, 3, 3, 1, 3, 2, 3, 5, 4, 5, 4, 6, 5, 4, 5, 6, 6, 4))
plot(G)
g_results <- matrix(page_rank(G)$vector)
```

The `igraph` package handles decay using a damping factor of 0.85 and automatically assigns a uniform random probability to dangling nodes, which matches the two approaches outlined above.  The page rank vector returned matches that returned through power iteration:

```{r comp-results}
identical(round(it_results, 13), round(g_results, 13))
```


### Comparison of Results
As shown in the above sections, the page rank vector for the given universe of six pages was derived through three methods:

  * Power iteration of the matrix $B$
  * Eigenvector corresponding to $\lambda = 1$ for the matrix $B$
  * `igraph` implementation using the matrix $A$
  
The three methods return the same results (to 13 decimal points of accuracy) once the eigenvector is scaled from being a unit vector.

```{r compare-results, echo=FALSE}
all_results <- data.frame(
  "Power Iteration" = round(ev_results, 13), 
  Eigenvector = round(ev_results, 13), 
  Graph = round(g_results, 13), 
  Rank = rank(1 - ev_results),
  row.names = seq(1, 6), 
  check.names = FALSE)

kable(all_results, padding = 0, align = 'c', row.names = TRUE)
```
