---
title: 'DATA 605 Assignment 11: Linear Regression'
author: "Dan Smilowitz"
date: "November 4, 2016"
output: 
  html_document: 
    highlight: tango
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = NA, fig.align = 'center')
library(ggplot2)
library(pander)
```



## Heart Rate Data

```{r heart}
hr <- data.frame(
  Age = c(18, 23, 25, 35, 65, 54, 34, 56, 72, 19, 23, 42, 18, 39, 37),
  MaxHR = c(202, 186, 187, 180, 156, 169, 174, 172, 153, 199, 193, 174, 198, 183, 178)
)

heart_fit <- lm(MaxHR ~ Age, data = hr)
```

`r pander(heart_fit)`

The equation fit to the data is $$\widehat{MaxHR} = `r round(heart_fit$coefficients[1], 3)` `r round(heart_fit$coefficients[2], 3)` \times age$$

The $p$ value for the coefficient `r round(heart_fit$coefficients[2], 3)` is $\approx 4 \times 10^{-8}$, showing that it is significant at a significance level of $\alpha = 0.01$.

```{r plot-heart, echo=FALSE}
heart_eq = paste0('MaxHR = ', 
                  round(heart_fit$coefficients[1], 3), ' - ', 
                  abs(round(heart_fit$coefficients[2], 3)), ' * age')
ggplot(hr, aes(Age, MaxHR)) + geom_point() + 
  geom_abline(slope = heart_fit$coefficients[2], intercept = heart_fit$coefficients[1]) + 
  ggtitle('Max Heart Rate vs. Age') + annotate('text', 60, 185, label=heart_eq)
```



## Auto Data

```{r auto-data}
auto <- read.table('data/auto-mpg.data')
names(auto) <- c('displacement', 'horsepower', 'weight', 'acceleration', 'mpg')
```


### Subset of Data
```{r auto-sub}
set.seed(42) # set seed for replicable results
auto_sub <- auto[sample(1:nrow(auto), 40), ]
sub_fit <- lm(mpg ~ ., auto_sub)
```

`r pander(sub_fit)`

The equation fit to the data is $$\widehat{mpg} = `r round(sub_fit$coefficients[1], 2)` `r round(sub_fit$coefficients[2], 2)` \times displacement + `r round(sub_fit$coefficients[3], 2)` \times horsepower + `r round(sub_fit$coefficients[4], 2)` \times weight + `r round(sub_fit$coefficients[5], 2)` \times acceleration$$

None of the coefficients are statistically significant at the $\alpha = 0.10$ significance level.  The standard errors for the coefficients are presented in the table above.  95% confidence intervals for the coefficients are as follows:

`r pander(confint(sub_fit))`


### Full Dataset
```{r auto-full}
auto_fit <- lm(mpg ~ ., auto)
```

`r pander(auto_fit)`

The equation fit to the data is $$\widehat{mpg} = `r round(auto_fit$coefficients[1], 2)` `r round(auto_fit$coefficients[2], 2)` \times displacement + `r round(auto_fit$coefficients[3], 2)` \times horsepower + `r round(auto_fit$coefficients[4], 2)` \times weight + `r round(auto_fit$coefficients[5], 2)` \times acceleration$$

The coefficients for `horsepower` and `weight` are statistically significant at the $\alpha = 0.01$ significance level; the coefficients for the remaining variables are not significant.  The standard errors for the coefficients are presented in the table above.  95% confidence intervals for the coefficients are as follows:

`r pander(confint(auto_fit))`
