---
title: 'DATA 605 Assignment 9: Central Limit Theorem'
author: "Dan Smilowitz"
date: "October 27, 2016"
output: 
  html_document: 
    highlight: tango
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = NA, fig.align = 'center')
```


## Sampling Functions

### Equation 1
$$\begin{align}
&f(x) = x&,\  0 \leq x \leq 1 \\
&f(x) = 2 - x&,\  1 \lt x \leq 2
\end{align}$$

This equation represents a triangular function from 0 to 2, with a maximum at 1, as shown below.
```{r plot-f1, echo=FALSE}
x1 <- seq(0, 1, 0.01); x2 <- seq(1, 2, 0.01)
y1 <- x1; y2 <- 2 - x2
x <- c(x1, x2)
y <- c(y1, y2)
plot(x, y, 'l', main = NULL)
```

This can be sampled from using the `triangle` package:
```{r sample-f1}
library(triangle)
sample_f1 <- function(N) {
  f1_samp <- rtriangle(N, a = 0, b = 2)
  f1_samp
}
```

### Equation 2
$$\begin{align}
&f(x) = 1 - x&,\  0 \leq x \leq 1 \\
&f(x) = x - 1&,\  1 \lt x \leq 2
\end{align}$$

This equation represents two triangular distributions, as shown below:

```{r plot-f2, echo=FALSE}
y1 <- 1 - x1; y2 <- x2 - 1
y <- c(y1, y2)
plot(x, y, 'l', main = NULL)
```

To sample the two triangular distributions, each is sampled, the two samples are combined.
```{r sample-f2}
sample_f2 <- function(N) {
  f2_samp_lower <- rtriangle(N/2, a = 0, b = 1, c = 0)
  f2_samp_upper <- rtriangle(N/2, a = 1, b = 2, c = 2)
  f2_samp = c(f2_samp_lower, f2_samp_upper)
  f2_samp
}
```

## Drawing Samples
Each of the samples is drawn 1000 times, and the histograms of the values in the sample are plotted in a histogram:

```{r sample-1000, fig.width=8}
my_f1 <- sample_f1(1000)
my_f2 <- sample_f2(1000)
par(mfrow = c(1, 2))
hist(my_f1, main = 'Sample from Equation 1')
hist(my_f2, main = 'Sample from Equation 2')
```

The resulting histograms roughly map the shape of the plots shown above.

## $n$ Samples
The function below performs $n$ samples of 1000 iterations from a distribution and returns plots a histograms of the $n$ sample means.
```{r sample-means}
central_limit <- function(n, PDF) {
  xbar <- NULL
  for (i in 1:n) {
    samp <- do.call(PDF, list(1000))
    xbar <- c(xbar, mean(samp))
  }
  plot_title <- paste(n, 'samples:', as.character(substitute(PDF)))
  hist(xbar, xlab = NULL, ylab = NULL, main = plot_title)
}
```

## Visualizing Central Limit Theorem
Sampling each distribution 5, 10, 20, 100, and 500 times produces the following results:
```{r CLT, fig.height=10, fig.width=8}
ns <- c(5, 10, 20, 100, 500)
par(mfrow = c(5, 2), mar = c(3,3,3,3))
for (i in 1:5) {
  central_limit(ns[i], sample_f1)
  central_limit(ns[i], sample_f2)
}
```

The plots above show that as the number of samples drawn increases, the sample means approaches a random distribution centered near the center of the PDF from which they are sampling.
